---
title: 'Select and rename columns'
order: 1
---

In this tutorial we use the `tidyverse` and our simulated [practice data](../datasets/practice-data.csv).

Open the tidyverse package, and read the data from the URL.

```{r, eval=T}
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
d <- read_csv(url)
```

## Selecting certain columns

Where `filter` selects specific rows, `select` allows you to select specific columns. 
Most simply, we can simply name the columns that we want to retrieve them in that particular order. 

```{r}
select(d, Population, Support, Pollster)
```

You can also specify a range of columns, for example all columns from Support to Democratic Support:

```{r}
select(d, Support:`Democratic Support`)
```

Note the use of 'backticks' (reverse quotes) to specify the column name, as R does not normally allow spaces in names. 

Select can also be used to rename columns when selecting them, for example to get rid of the spaces:

```{r}
select(d, Pollster, rep = `Republican Support`, dem = `Democratic Support`)
```

Note that `select` drops all columns not selected. If you only want to rename columns, you can use the `rename` function:

```{r}
rename(d, start_date = Start, end_date = End)
```

Finally, you can drop a variable by adding a minus sign in front of a name:

```{r}
select(d, -Question, -URL)
```


