---
title: 'Select and rename'
order: 1

---

::: {.callout-note title="Required packages and data for this tutorial"}

In this tutorial we use the `tidyverse` package and the simulated [practice data](../../data-management/index.qmd#practice-data).

```{r, message=FALSE}
library(tidyverse)
d <- read_csv("https://tinyurl.com/R-practice-data")
```
:::

# Selecting columns with `select`

Often you do not need to use all columns in your data, or you only need a subset of the columns for a specific analysis.
You can do this with the `select` function.

First, let's see what columns are in our data using the `colnames` function, which returns the column names of a data frame:

```{r}
colnames(d)
```

### Selecting specific columns

The simplest way of using `select` is to explicitly specify the columns you want to keep:

```{r, results='hide'}
ds <- select(d, age, np_subscription)
```

This will return a new tibble with only the columns `id`, `age`, and `np_subscription`.
We assigned this new tibble to the variable `ds` (short for "data subset").
Sometimes you want to overwrite the original data frame with the new selection.
You can do this by assigning the result to the same name as the input (`d` in this case):

```{r, results='hide', eval=FALSE}
d <- select(d, age, np_subscription)
```

Just be carefull with this. 
One of the nice things about R is that you can have multiple versions of your data in different tibbles.
It is often smart to at least keep the original (raw) data frame intact.

### Selecting a range of columns

You can also specify a range of columns using the syntax `first_column:last_column`.
For example, to select all columns from `experiment_group` to `trust_t2`:

```{r, results='hide'}
select(d, experiment_group:trust_t2)
```

This will return a new tibble with only the columns `experiment_group`, `trust_t1`, and `trust_t2`.

Note that here we did not assign the result to anything.
So in this case R will just print the result to the console, but not store it in a variable.

### Selecting and renaming columns

When you select a column, you can also rename it using the syntax `new_name = old_name`. 
The following code selects the columns `experiment_group`, `trust_t1`, and `trust_t2`, and renames them to `group`, `trust_before`, and `trust_after`:

```{r, results='hide'}
select(d, group = experiment_group, 
          trust_before = trust_t1, 
          trust_after = trust_t2)
```


### Selecting columns that have spaces in the name  

Sometimes columns names have spaces in them.
This is a bit annoying to work with in R, because you need to then tell R where a name starts and ends.
You can do this by using backticks (reverse quotes) around the column name. 
In our practice data, we need this to select the `news consumption` column.
It is then often smart to immediately rename the column to something without spaces, such as just replacing them with underscores:

```{r, results='hide'}
select(d, news_consumption = `news consumption`)
```

### Dropping columns

Instead of selecting which column to keep, you can also specify which columns to drop.
You can do this by adding a minus sign in front of the column name.
The following code drops the columns `np_subscription` and `trust_t1`:

```{r, results='hide'}
select(d, -np_subscription, -trust_t1)
```

This will return a new tibble with all columns except `np_subscription` and `trust_t1`.


# Renaming columns with `rename`

Sometimes you only want to rename columns without selecting or dropping any.
You can do this with the `rename` function, which works similarly to how you rename columns with `select`:

```{r, results='hide'}
rename(d, group = experiment_group, 
          trust_before = trust_t1, 
          trust_after = trust_t2)
```

In this case, we do rename the columns, but without dropping all the other columns.
