---
title: Packages
order: 5
---

# What is a package? {#what-is-a-package}


In R, a package is a collection of functions, data, and documentation that extends the capabilities of R. 
You can think of packages kind of like apps on your phone: they provide additional functionality that you can use to perform specific tasks.
Also, like apps, you can install and uninstall packages as needed, directly from within R.

There are thousands of R packages available, which enables you to use almost any existing data analysis technique.
R is not just a tool for statistical analysis, but also for data visualization, data collection, articifial intelligence, and much more.
If there is anything you need to do, there is a good chance that someone has already written a package for it.

# How to install and use a package

To use a package, you first need to install it. **Installing a package only needs to be done once** per system (unless you need to update it).

## 1. Install a package

Most packages are available on the Comprehensive R Archive Network (CRAN), which is the main repository for R packages.
To install these packages, all you need to know is their name.

For example, there is a package called `lubridate` that makes it easier to work with dates and times in your data.
To install this package, you can use the `install.packages()` function:

```{r, eval=F}
install.packages("lubridate")
```

## 2. Load a package

Once you have installed a package, it is not yet loaded into your current R session.
Similar to when you install a new app on your phone, you need to open it every time you want to use it.

To use a package in your current R session, you can load it with the `library()` function:

```{r, message=FALSE}
library(lubridate)
```

# How do I know how to use a package?

Most packages come with good documentation that explains how to use the functions in the package.
For individual functions you can use the `?` operator to open the help page for that function, as we explain in [here](../good-to-know/function-documentation.qmd).
But there is often also a more general documentation that explains the package as a whole.
This is called a **vignette**, and if a package has one, you can open it with the `vignette()` function.

```{r, eval=F}
vignette("lubridate")
```

You can look up vignettes and function documentation [online](https://www.rdocumentation.org/). 
This is also a good way to find new packages for whatever you're trying to do.

