
```{r}
library(tidyverse)
d = read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv")
```


## Viewing a summary of every column

Often, the main information that you're interested in is the summary of each column.
You can compute this yourself (as we'll show later), but there is also a nice package called `summarytools` that can generate a handy overview.
You can create the summary as follows (remember to install the package first):

```{r, output=FALSE, message=FALSE}
library(summarytools)
dfSummary(d)
```

This will print the summary in your console, but you can also render it in a more readable format with the `view` function.

```{r, eval=FALSE}
view(dfSummary(d))
```

```{r, echo=FALSE}
print(dfSummary(d), 
      method='render', 
      max.tbl.height = 500)
```

::: {.callout-note title="Using summarytools in Quarty or RMarkdown" collapse="true"}

If you're working in Quarto or RMarkdown, the normal way of using `summarytools` doesn't create a nice table. Instead, what you need to do is tell Quarto to render the output as HTML. You can do this by setting the `method` argument to `render`.

```{r, eval=FALSE}
print(dfSummary(iris), 
      method='render', 
      max.tbl.height = 500)
```

Note that we also set the `max.tbl.height` argument to 500, to make sure that very large tables are put inside a scrollable window.
:::
